---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# socialchange

<!-- badges: start -->
[![R-CMD-check](https://github.com/elbersb/socialchange/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/elbersb/socialchange/actions/workflows/R-CMD-check.yaml)
[![Codecov](https://codecov.io/gh/elbersb/socialchange/branch/main/graph/badge.svg)](https://app.codecov.io/gh/elbersb/socialchange)
[![pkgdown](https://github.com/elbersb/socialchange/actions/workflows/pkgdown.yaml/badge.svg)](https://elbersb.github.io/socialchange/)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The `socialchange` package provides methods for decomposing aggregate-level social change
into components of intraindividual change and population turnover. Population
turnover includes replacement of individuals (mortality and coming-of-age) and migration.
The package implements event-based, aggregated-data, and simulation-based decomposition
approaches, as well as classical cohort replacement vs. intraindividual change
(CR-IC) decompositions and Age-Period-Cohort (APC) models.

## Installation

You can install the development version of socialchange from [GitHub](https://github.com/elbersb/socialchange) with:

``` r
# install.packages("pak")
pak::pak("elbersb/socialchange")
```

## Example: Direct decomposition of events

As a basic example, we look at a decomposition of EU membership, and whether
population growth in the EU is due to intracountry changes in population, or due to
the admission of new countries. The decomposition unit is the country.
To run the `decompose_events` decomposition function, we need two datasets:
One dataset contains the events, in this case the dates when countries entered
or exited the EU. This dataset is included in the `socialchange` package:


```{r}
library(socialchange)
socialchange::eu_membership
```

The second dataset contains the outcome that we want to decompose
at any point in time where there are change events, plus the start and end period of interest.
The EU was formed in 1952, and then countries entered or exited in 1973, 1981, 1986, 1995, 2004, 2007, 2013, and 2020.
Countries, of course, entered on specific dates, but most population data is only provided at yearly resolutions, so
we assume that the error introduced by using a yearly resolution is small. The `socialchange` package also includes
total population data for all countries of the world from 1950 onwards, so it also includes the required data:

```{r}
library(data.table)
wpp_data <- as.data.table(socialchange::wpp_data)
# population data is in 1,000
wpp_data[Location == "Belgium" & Time %in% c(1952, 1973, 1981, 1986, 1995, 2004, 2007, 2013, 2020)]
```

To run the `decompose_events` function, we need to ensure that columns for the unit and time parameters
are named the same, so we prepare the data as follows:

```{r example}
# rename columns
setnames(wpp_data, c("Location", "Time"), c("country", "year"))
# population in millions
wpp_data[, PopTotal := PopTotal / 1000]

# coarsen to year - required as we don't have daily population data
eu_membership <- as.data.table(socialchange::eu_membership)
eu_membership[, year := year(as.Date(date))]
```

Finally, we run the decomposition. We specify the two datasets, and a formula of the form `Outcome ~ Unit + Time`.
We also need to specify the end data, and an aggregation function. Often this will be mean, but in our case it needs
to be `sum`:

```{r}
ev <- decompose_events(eu_membership, wpp_data, PopTotal ~ country + year,
  end_period = 2021, fun = sum)
print(ev)
```

The outcome of the decomposition tells us that intracountry changes in population
account for 30% of the total population increase (or 80 million), while additions and one (Br)exit
account for 70% of the total population increase (or 184 million).

It is also possible to produce a figure that shows the decomposition:

```{r}
plot(ev)
```

## Example: Decompose aggregated data

In this example we decompose survey data, based on the U.S. General Social Survey.
More details on this example can be found in the [vignette](https://elbersb.github.io/socialchange/articles/gss_homosexuality.html#event-based-decomposition).

```{r}
data(gss_homosex)

gss_all <- gss_homosex[age >= 21, .(age, period = year, sex, y = homosex, wtssall)]
model <- mgcv::gam(y ~ s(age) + s(period), data = gss_all, weights = wtssall)
result <- decompose_aggregated(gss_all, model, cells = "sex", weight = "wtssall", R = 100)
print(result, detailed = FALSE)
plot(result)
```
