---
title: "Decomposing General Social Survey data: Attitudes on Homosexuality"
vignette: >
  %\VignetteIndexEntry{Decomposing General Social Survey data: Attitudes on Homosexuality}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

The package includes a dataset `gss_homosex`, covering General Social Survey data on attitudes
towards homosexuality in the U.S. This data was also analyzed in Ekstam (2021).

## Packages

```{r}
library("socialchange")
library("modelsummary")
library("ggplot2")
library("mgcv")

data(gss_homosex)
```

## Descriptives

The `homosex` outcome is rescaled to [0, 1], where 0 means "always wrong" and 1 means "not wrong at all". The table below gives summary statistics; the line plot shows a clear upward trend in acceptance over the survey period.

```{r}
# compare to Table 1 in Ekstam -- roughly similar
modelsummary::datasummary(
  homosex + year + cohort + age + educ + sex + race ~ mean + SD + min + max,
  data = gss_homosex, output = "markdown", fmt = 3)

by_year <- gss_homosex[, list(y = weighted.mean(homosex, wtssall)), by = c("year")]
ggplot(by_year, aes(x = year, y = y)) + geom_line() + ylim(0, 1) + theme_light()
```

To separate period and cohort effects, we fit a GAM with a two-dimensional smooth over survey year and birth cohort. The 3D surface and the contour plot below both shows that acceptance increased more strongly among younger cohorts and in later periods.

```{r}
splinemodel = gam(homosex ~ s(year, cohort), data = gss_homosex)
vis.gam(splinemodel, theta = 40)
plot_gam_surface(splinemodel)
```

## CR-IC decomposition

The `cr_ic()` function implements four methods for separating intracohort change (IC) from cohort replacement (CR): algebraic decomposition (AD), linear decomposition (LD), and two model-based improvements (AD+ and Model). For a detailed explanation of each method, including a replication of Firebaugh (1989), see the [Replications: Firebaugh](replicating_firebaugh.html) vignette.

Acceptance rose by about 43 percentage points between 1973 and 2018 (from 0.19 to 0.62). Using the full panel, the model-based methods attribute roughly 56% of this change to intracohort change and 44% to cohort replacement, meaning attitude shifts within individuals account for the larger share.

```{r}
# complete period
form <- homosex ~ as.factor(year) + as.factor(cohort)
(res <- cr_ic(gss_homosex, homosex ~ year + cohort, weight = "wtssall", model = form))
```

```{r}
plot(res)
```

When only the first and last survey years are used, the balance shifts slightly toward cohort replacement (~54% CR, ~46% IC). This illustrates how the choice of time points can affect decomposition results, and why the model-based AD+ and Model methods are preferred over the simpler AD.

```{r}
# only beginning and end year (also compare to Baunach 2011)
form <- homosex ~ as.factor(year) + splines::bs(cohort, 10)
(res <- cr_ic(gss_homosex[year %in% c(min(year), max(year))], homosex ~ year + cohort, weight = "wtssall", model = form))
```

For APC analysis of the same `gss_homosex` dataset, see the [APC vignette](apc.html).
