---
title: "Decomposing General Social Survey data: Attitudes on Homosexuality"
vignette: >
  %\VignetteIndexEntry{Decomposing General Social Survey data: Attitudes on Homosexuality}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

The package includes `gss_homosex`, covering General Social Survey data on attitudes
towards homosexuality in the U.S. from 1973 through 2024. A similar analysis of earlier
waves was done by Ekstam (2021).

## Packages

```{r}
#| message: false
library("socialchange")
library("modelsummary")
library("ggplot2")
library("mgcv")

data(gss_homosex)
```

## Descriptives

The `homosex` outcome is rescaled to [0, 1], where 0 means "always wrong" and 1 means "not wrong at all". The question was not asked in 1972, 1975, 1978, 1983, or 1986, so those years are absent. In included years, rows without a valid response are excluded. Most such rows reflect planned ballot assignment; the remainder are item nonresponse. The table below gives summary statistics; the line plot shows the trend in acceptance over the survey period.

```{r}
modelsummary::datasummary(
  homosex + year + cohort + age + educ ~ mean + SD + min + max,
  data = gss_homosex, output = "markdown", fmt = 3)

# sex and race are categorical, so show their level breakdown instead
modelsummary::datasummary(
  sex + race ~ N + Percent(),
  data = gss_homosex, output = "markdown", fmt = 1)

by_year <- gss_homosex[, list(y = weighted.mean(homosex, wtssps)), by = c("year")]
ggplot(by_year, aes(x = year, y = y)) + geom_line() + ylim(0, 1) + theme_light()
```

`wtssps` is the GSS post-stratification weight recommended for analyses spanning the full series. It incorporates the 1982 and 1987 Black oversamples, so those respondents remain in the data. Data collection changed after 2018: the 2021 round had no in-person interviews, and later rounds used mixed modes. Changes in the final waves can therefore reflect mode effects as well as attitude change.

To separate period and cohort effects, we fit a GAM with a two-dimensional smooth over survey year and birth cohort. The 3D surface and the contour plot below both show that acceptance increased more strongly among younger cohorts and in later periods.

```{r}
splinemodel = gam(homosex ~ s(year, cohort), data = gss_homosex)
vis.gam(splinemodel, theta = 40)
plot_gam_surface(splinemodel)
```

## CR-IC decomposition

The `cr_ic()` function implements four methods for separating intracohort change (IC) from cohort replacement (CR): algebraic decomposition (AD), linear decomposition (LD), and two model-based improvements (AD+ and Model). For a detailed explanation of each method, including a replication of Firebaugh (1989), see the [Replications: Firebaugh](replicating_firebaugh.html) vignette.

Acceptance rose by about 42 percentage points between 1973 and 2024 (from 0.20 to 0.61). Using all waves, AD+ attributes about 60% to intracohort change and 40% to cohort replacement, while the Model method gives 46% and 54%, respectively.

```{r}
# complete period
form <- homosex ~ as.factor(year) + as.factor(cohort)
(res <- cr_ic(gss_homosex, homosex ~ year + cohort, weight = "wtssps", model = form))
```

```{r}
plot(res)
```

When only the first and last survey years are used, AD+ attributes about 59% to cohort replacement and 41% to intracohort change; the Model method gives 55% and 45%. This illustrates how the choice of time points can affect the decomposition and why the model-based methods are preferred over the simpler AD.

```{r}
# only beginning and end year (also compare to Baunach 2011)
form <- homosex ~ as.factor(year) + splines::bs(cohort, 10)
(res <- cr_ic(gss_homosex[year %in% c(min(year), max(year))], homosex ~ year + cohort, weight = "wtssps", model = form))
```

## Event-based decomposition

The `decompose_aggregated()` function separates change in the aggregate outcome into intraindividual change (attitude shifts within respondents already in the population) and population turnover (mortality and coming-of-age effects). We rename `year` and `homosex` to the column names the function expects, then pass the full dataset directly.

```{r}
# All waves span ages 18--89+. Pool ages 81 and older into a stable open
# outcome cell.
gss_all <- gss_homosex[,
    .(age = pmin(age, 81), period = year, sex, y = homosex, wtssps)]
```

We fit a GAM to the individual responses to model acceptance as a smooth function of age and period.

```{r}
set.seed(42)
model <- mgcv::gam(y ~ s(age) + s(period), data = gss_all, weights = wtssps)
```

### The simplest decomposition

We can run the simplest case of the event decomposition with the following commmand:

```{r}
result <- decompose_aggregated(gss_all, model, weight = "wtssps")
print(result, detailed = FALSE)
```

This decomposition compares cells across years, attributing shrinking cohorts to mortality and the youngest cohorts to coming-of-age. Every change in a cell's size must be attributed to *some* demographic event — the decomposition never leaves a residual — so a survivor cohort that *grows* between two waves is credited to net in-migration. With only the survey to work from, this in-migration contribution (about 0.057, or 13% of total change) largely reflects sampling fluctuation in cell sizes across waves rather than genuine immigration. It becomes meaningful only once a true population frame is supplied (see below).

Of the roughly 43 percentage point modeled rise between 1973 and 2024, the decomposition attributes about 58% to intraindividual change and 42% to population turnover. This turnover share is within the 41--54% range from the CR-IC model-based methods above.

```{r}
plot(result)
```

The components also differ over time. Intraindividual change was negative through much of the 1970s and 1980s, while turnover was positive. From the early 1990s through 2018, intraindividual change became the main source of rising acceptance. It turned negative again in 2022 and 2024, alongside the decline in the observed mean after 2021. The survey-design changes in these final waves prevent a purely attitudinal interpretation of that reversal.

One speculative reading of the early decline in intraindividual acceptance is the backlash associated with the AIDS epidemic of the 1980s. The decomposition itself cannot establish this, but it does isolate a within-person decline that is invisible in the raw trend line.

### Adding a covariate

We can split the population into covariate cells with the `cells` argument. Adding `sex` lets the outcome model predict differently for men and women, and — more consequentially for the decomposition — derives the demographic events (mortality, coming-of-age, migration) separately within each sex cell rather than from the pooled age structure.

```{r}
model <- mgcv::gam(y ~ s(age) + s(period) + sex, data = gss_all, weights = wtssps)
set.seed(42)
result <- decompose_aggregated(gss_all, model, cells = "sex", weight = "wtssps")
print(result, detailed = FALSE)
plot(result)
```

The headline split is essentially unchanged from the pooled model: about 58% intraindividual change and 42% turnover. Conditioning on sex barely moves the population-weighted mean because the sex composition is roughly stable across waves.

What the covariate also adds is an *attribution* of every component to the cells that produced it. Passing `covariate = "sex"` to `print()` or `plot()` splits each row into the contribution of each sex to the overall change. The split is fully additive, so this is a genuine decomposition of the aggregate change.

```{r}
print(result, detailed = FALSE, covariate = "sex")
plot(result, covariate = "sex")
```

Women and men contribute similar amounts to total change (0.217 and 0.212), intraindividual change (0.121 and 0.126), and turnover (0.096 and 0.086).

The structure within turnover needs a caveat. Mortality contributes −0.036 for women and 0.115 for men, offset by in-migration contributions of 0.102 and −0.050. This asymmetry should not be read demographically. On survey data, each survivor cell's net size change is routed entirely to mortality if it shrank or to in-migration if it grew. Splitting noisy survey cells by sex amplifies sampling fluctuation, so this division is largely an artifact of survey cell sizes. A population frame improves the attribution.

### Adding a population frame

So far the cell counts `n` have come from the survey itself, so the demographic events are inferred from the noisy survey cell sizes across waves. When true population counts are available, the `population` argument lets the survey supply only the outcome `model`, while an external frame supplies the cell counts that drive the demographic events.

The package ships `wpp_us`, US population by age and sex from the UN World Population Prospects for each survey year. Only the *relative* cell structure matters, so we rescale each wave to a tractable per-period total before passing it in.

```{r}
data(wpp_us)

survey_years <- sort(unique(gss_all$period))
pop <- wpp_us[period %in% survey_years]
# Keep ages 81--88 and the terminal 89+ group separate for demographic
# calculations. Outcome predictions for all these ages use the survey's 81+ cell.
scale_total <- round(mean(gss_all[, .N, by = period]$N))
pop[, n := n / sum(n) * scale_total, by = period]

set.seed(42)
result <- decompose_aggregated(gss_all, model, cells = "sex",
                               weight = "wtssps", population = pop)
print(result, detailed = FALSE)
plot(result)
```

The population frame gives about 59% intraindividual change and 41% turnover. Its smooth age structure removes survey cell-size noise: mortality contributes 0.084, coming-of-age contributes 0.088, and net in-migration falls to 0.004.

The clearest payoff shows up in the per-sex breakdown, which we can compare directly with the survey-counts version above.

```{r}
print(result, detailed = FALSE, covariate = "sex")
plot(result, covariate = "sex")
```

Mortality now splits into demographically plausible contributions of 0.033 for women and 0.051 for men. The large offsetting in-migration terms collapse to 0.003 and 0.001. This confirms that the earlier mortality/migration pattern came from small survey cells.

### Adding mortality data

Even with a population frame, the split of each survivor cell's change between mortality and migration rests on a sign heuristic: per cell the balancing identity is $n_2 = n_1 - \text{deaths} + \text{net migration}$ — one equation, two unknowns — and the default closes it by routing the whole net change to mortality if the cell shrank and to in-migration if it grew. External mortality data closes the identity directly: deaths come from real death probabilities, and migration becomes the *signed residual*, so out-migration can appear for the first time.

The package ships `mortality_us`, HMD central death rates $m(x)$ by year, age, and sex. The `mortality` argument expects annual death *probabilities* $q(x)$, so we convert with $q = 1 - e^{-m}$ and rename `year` to `period`; the `sex` column matches `cells = "sex"`. Both WPP and HMD end in a matching 89+ open group. The decomposition retains WPP's true ages 81--89 for mortality, while clamping their outcome predictions to the survey's 81+ cell.

```{r}
data(mortality_us)
mort <- mortality_us[, .(period = year, age, sex, prob = 1 - exp(-death_rate))]

set.seed(42)
result <- decompose_aggregated(gss_all, model, cells = "sex",
                               weight = "wtssps", population = pop,
                               mortality = mort)
print(result, detailed = FALSE)
plot(result)
```

The headline split is stable, but annual death timing causes small compensating changes across path-dependent components. Mortality falls from 0.084 under sign attribution to 0.053 with HMD probabilities, while a new out-migration term contributes 0.033. The WPP cohorts shrink across waves by more than HMD survival accounts for, so residual migration records the difference instead of folding all shrinkage into mortality.

That out-migration term is not a direct migration estimate. It combines genuine net migration with inconsistency between WPP cohort counts and HMD survival. About 0.008 of its 0.033 contribution sits in the survey's 81+ outcome cell, where deaths are calculated from WPP's constituent ages 81--89. With that qualification, supplied mortality probabilities attribute about 0.053, or 12% of total attitude change, to mortality itself.

```{r}
print(result, detailed = FALSE, covariate = "sex")
```

The mortality contribution is nearly even by sex (0.025 for women and 0.028 for men). Out-migration contributes more among men (0.024 versus 0.009), but this remains a residual difference rather than an identified sex difference in emigration.

### Standard errors

Passing `R > 0` attaches standard errors and 95% confidence bands to every component. Each replicate re-runs the decomposition under a fresh random ordering of the interleaved demographic events and a refit of the outcome `model` on a reweighted bootstrap sample of the survey; the spread across the `R` replicates is the band that `print()` reports and `plot()` shades.

```{r}
set.seed(42)
result <- decompose_aggregated(gss_all, model, cells = "sex",
                               weight = "wtssps", population = pop, R = 100,
                               seed = 42)
print(result, detailed = FALSE)
plot(result)
```

The bands separate the roughly 59% intraindividual and 41% turnover contributions: both are well above zero, and their intervals do not overlap. Intraindividual change and coming-of-age have the widest intervals; mortality and the small net in-migration term are estimated more tightly.

Second, the bands combine model uncertainty with event-ordering uncertainty only. Because the population frame supplies the cell counts, those counts are treated as fixed, so the intervals do not include demographic sampling error in the turnover the counts drive — they would understate total uncertainty if the cell sizes were themselves noisy (as they are when the counts come from the survey, for instance, rather than a population frame).

Standard errors can also be obtained when we split by covariate:

```{r}
print(result, detailed = FALSE, covariate = "sex")
plot(result, covariate = "sex")
```

The female and male intervals overlap for the total change and for every component except mortality (and the tiny net in-migration term), so the apparent differences in intraindividual change and coming-of-age are within sampling noise rather than distinguishable differences. Mortality is the exception: the male contribution is clearly larger, in line with men's excess mortality.

## Further reading

For APC analysis of the same `gss_homosex` dataset, see the [APC vignette](apc.html).
