---
title: "Decomposing General Social Survey data: Attitudes on Homosexuality"
vignette: >
  %\VignetteIndexEntry{Decomposing General Social Survey data: Attitudes on Homosexuality}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

The package includes a dataset `gss_homosex`, covering General Social Survey data on attitudes
towards homosexuality in the U.S. This data was also analyzed in Ekstam (2021).

## Packages

```{r}
library("socialchange")
library("modelsummary")
library("ggplot2")
library("mgcv")

data(gss_homosex)
```

## Descriptives

The `homosex` outcome is rescaled to [0, 1], where 0 means "always wrong" and 1 means "not wrong at all". The table below gives summary statistics; the line plot shows a clear upward trend in acceptance over the survey period.

```{r}
# compare to Table 1 in Ekstam -- roughly similar
modelsummary::datasummary(
  homosex + year + cohort + age + educ + sex + race ~ mean + SD + min + max,
  data = gss_homosex, output = "markdown", fmt = 3)

by_year <- gss_homosex[, list(y = weighted.mean(homosex, wtssall)), by = c("year")]
ggplot(by_year, aes(x = year, y = y)) + geom_line() + ylim(0, 1) + theme_light()
```

To separate period and cohort effects, we fit a GAM with a two-dimensional smooth over survey year and birth cohort. The 3D surface and the contour plot below both shows that acceptance increased more strongly among younger cohorts and in later periods.

```{r}
splinemodel = gam(homosex ~ s(year, cohort), data = gss_homosex)
vis.gam(splinemodel, theta = 40)
plot_gam_surface(splinemodel)
```

## CR-IC decomposition

The `cr_ic()` function implements four methods for separating intracohort change (IC) from cohort replacement (CR): algebraic decomposition (AD), linear decomposition (LD), and two model-based improvements (AD+ and Model). For a detailed explanation of each method, including a replication of Firebaugh (1989), see the [Replications: Firebaugh](replicating_firebaugh.html) vignette.

Acceptance rose by about 43 percentage points between 1973 and 2018 (from 0.19 to 0.62). Using the full panel, the two model-based methods both attribute the larger share to intracohort change, though they differ on the exact split: AD+ gives roughly 63% intracohort change and 37% cohort replacement, while the Model method gives roughly 56% and 44%.

```{r}
# complete period
form <- homosex ~ as.factor(year) + as.factor(cohort)
(res <- cr_ic(gss_homosex, homosex ~ year + cohort, weight = "wtssall", model = form))
```

```{r}
plot(res)
```

When only the first and last survey years are used, the balance shifts slightly toward cohort replacement (~54% CR, ~46% IC). This illustrates how the choice of time points can affect decomposition results, and why the model-based AD+ and Model methods are preferred over the simpler AD.

```{r}
# only beginning and end year (also compare to Baunach 2011)
form <- homosex ~ as.factor(year) + splines::bs(cohort, 10)
(res <- cr_ic(gss_homosex[year %in% c(min(year), max(year))], homosex ~ year + cohort, weight = "wtssall", model = form))
```

## Event-based decomposition

The `decompose_aggregated()` function separates change in the aggregate outcome into intraindividual change (attitude shifts within respondents already in the population) and population turnover (mortality and coming-of-age effects). We rename `year` and `homosex` to the column names the function expects, then pass the full dataset directly.

```{r}
gss_all <- gss_homosex[, .(age, period = year, y = homosex, wtssall)]
```

We fit a GAM to the individual responses to model acceptance as a smooth function of age and period.

```{r}
set.seed(42)
model <- mgcv::gam(y ~ s(age) + s(period), data = gss_all, weights = wtssall)
predict_y <- function(newdata) as.vector(mgcv::predict.gam(model, newdata = newdata))

result <- decompose_aggregated(gss_all, predict_y, weight = "wtssall", tol = 0.15)
print(result)
```

Of the roughly 41 percentage point rise in acceptance between 1973 and 2018, the decomposition attributes about 61% to intraindividual change and 39% to population turnover (mortality of older, less accepting cohorts; coming-of-age of younger, more accepting ones). This falls within the range spanned by the CR-IC model-based methods above (37–44% turnover), closest to the AD+ estimate.

```{r}
plot(result)
```

What is most interesting about these results is how the components behave over time. Until the early-to-mid 1990s, intraindividual change and population turnover offset each other, leaving the aggregate mean roughly flat. The mean even dipped in the mid-1980s, falling from about 0.21 in 1977 to 0.18 in 1987, as the existing members of the population became less accepting of homosexuality. Turnover pulled the mean back up over the same period, so that by the early 1990s acceptance had barely moved from its 1973 level. Only from the mid-1990s did intraindividual change start to contribute toward acceptance, and from that point on it became the main driver of the rising trend. Turnover contributed in the positive direction throughout. Mortality of older, less accepting cohorts had a much stronger effect than the coming-of-age of younger ones, likely because new members of the population are already closer to the population mean.

One speculative reading of the early decline in intraindividual acceptance is the backlash associated with the AIDS epidemic of the 1980s. The decomposition itself cannot establish this, but it does isolate a within-person decline that is invisible in the raw trend line.

## Further reading

For APC analysis of the same `gss_homosex` dataset, see the [APC vignette](apc.html).
