---
title: "Decomposing General Social Survey data: Attitudes toward homosexuality"
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Decomposing General Social Survey data: Attitudes toward homosexuality}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

The package includes `gss_homosex`, which covers General Social Survey (GSS) data on attitudes toward homosexuality in the United States from 1973 through 2024. Ekstam [-@ekstam2021] analyzes the same GSS item through 2016. This vignette uses `decompose_aggregated()` to ask how much of the increase in acceptance reflects modeled change within the surviving population and how much reflects population turnover.

::: {.callout-tip title="Key findings"}
-   Acceptance increased by about 42 percentage points. The model attributes about 58% of this rise to modeled change within the surviving adult population (**intraindividual change**). The other 42% came from changes in who made up the population (**population turnover**): more accepting younger people entered adulthood, less accepting older people died, and the population changed through migration.
-   From 1973 through 1988, people already in the adult population became less accepting. Their opinion change pushed acceptance down, while the entry of younger people and the death of older people limited the decline. From 1988 through 2021, people already in the adult population became more accepting. This broad opinion change became the main source of rising acceptance, while generational replacement continued to add to it.
-   From 2021 through 2024, intraindividual change was negative while turnover remained positive. This provisional reversal forms a short third phase.
-   The broad split is stable across the demographic inputs used below. The allocation within turnover is not stable: population estimates and mortality probabilities improve the attribution of mortality and residual migration.
:::

## Packages

```{r}
#| message: false
library("socialchange")
library("modelsummary")
library("ggplot2")
library("mgcv")

data(gss_homosex)
```

## Observed change

The `homosex` outcome is rescaled to \[0, 1\], where 0 means "always wrong" and 1 means "not wrong at all." The question was not asked in 1972, 1975, 1978, 1983, or 1986, so those years are absent. In included years, rows without a valid response are excluded. Most exclusions reflect planned ballot assignment; the remainder are item nonresponse [@gss2024].

```{r}
modelsummary::datasummary(
  homosex + year + cohort + age + educ ~ mean + SD + min + max,
  data = gss_homosex, output = "markdown", fmt = 3)

# Sex and race are categorical, so show their level breakdown.
modelsummary::datasummary(
  sex + race ~ N + Percent(),
  data = gss_homosex, output = "markdown", fmt = 1)
```

`wtssps` is the GSS post-stratification weight recommended for analyses that span the full series. It incorporates the 1982 and 1987 Black oversamples, so those respondents remain in the data [@gss2024].

The weighted mean rose from about 0.20 in 1973 to 0.61 in 2024. The final waves need special care. The 2021 round had no in-person interviews, and later rounds used mixed modes. Their changes can reflect mode effects as well as attitude change [@gss2024].

```{r}
by_year <- gss_homosex[, .(acceptance = weighted.mean(homosex, wtssps)), by = year]

ggplot(by_year, aes(x = year, y = acceptance)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Acceptance of homosexuality increased substantially",
    subtitle = "Weighted General Social Survey mean",
    x = "Survey year", y = "Mean acceptance"
  ) +
  theme_light()
```

A smooth surface provides a descriptive view of period and cohort variation. Acceptance is higher in later periods and younger cohorts. This surface is not the outcome model used in the decomposition below and does not, by itself, identify causal period or cohort effects.

```{r}
splinemodel <- gam(
  homosex ~ s(year, cohort),
  data = gss_homosex,
  weights = wtssps
)
vis.gam(
  splinemodel,
  view = c("year", "cohort"),
  type = "response", theta = 40,
  ticktype = "detailed",
  xlab = "Survey year", ylab = "Birth cohort", zlab = "Acceptance",
  cex.axis = 0.8, cex.lab = 0.9
)
```

Ekstam [-@ekstam2021] describes higher tolerance among cohorts born from 1942 through 1951 than among adjacent cohorts, and lower tolerance among cohorts born from 1957 through 1966. He relates these descriptive patterns to formative exposure to the counterculture era and to the HIV/AIDS crisis and conservative climate of the 1980s. Neither this surface nor an APC model can identify that socialization explanation from age, period, and cohort patterns alone.

The figure below provides the closest direct comparison available in the GSS: weighted cohort means at ages 18--24. Because the series starts in 1973, it cannot show most of the 1942--1951 cohorts during young adulthood. The means also combine cohort and survey-period differences.

```{r}
#| echo: false
young_cohorts <- gss_homosex[
  age >= 18 & age <= 24,
  .(acceptance = weighted.mean(homosex, wtssps)),
  by = cohort
]
young_cohorts[, group := data.table::fifelse(
  cohort >= 1942 & cohort <= 1951, "1942--1951",
  data.table::fifelse(
    cohort >= 1957 & cohort <= 1966, "1957--1966", "Other cohorts"
  )
)]

ggplot(young_cohorts, aes(x = cohort, y = acceptance)) +
  annotate(
    "rect", xmin = 1941.5, xmax = 1951.5, ymin = -Inf, ymax = Inf,
    fill = "#0072B2", alpha = 0.08
  ) +
  annotate(
    "rect", xmin = 1956.5, xmax = 1966.5, ymin = -Inf, ymax = Inf,
    fill = "#D55E00", alpha = 0.08
  ) +
  geom_line(color = "grey65") +
  geom_point(aes(color = group), size = 2) +
  scale_color_manual(values = c(
    "1942-1951" = "#0072B2",
    "1957-1966" = "#D55E00",
    "Other cohorts" = "grey35"
  )) +
  coord_cartesian(xlim = c(1942, 2006), ylim = c(0, 1)) +
  labs(
    title = "Acceptance in young adulthood varies by birth cohort",
    subtitle = "Weighted GSS means among respondents aged 18-24",
    x = "Birth cohort", y = "Mean acceptance", color = "Birth cohort",
    caption = "The GSS starts in 1973; cohorts before 1949 are not observed at ages 18-24."
  ) +
  theme_light() +
  theme(legend.position = "bottom")
```

## What the decomposition measures

`decompose_aggregated()` represents aggregate change as the sum of two broad components:

| Component | Interpretation |
|------------------------------------|------------------------------------|
| Intraindividual change | Change in the modeled outcome among the surviving population as age and period advance |
| Population turnover | Change caused by mortality, coming-of-age, and net migration |

The method combines an outcome model with period-specific population cells. It places the demographic events between survey waves in random order and records each event's contribution to the population mean.

::: {.callout-important title="Interpretation"}
The GSS is a repeated cross-section. It does not follow the same respondents over time. "Intraindividual change" is therefore a model-based component, not a direct estimate of within-person change from panel observations. The plain-language descriptions below interpret this component as people changing. That plausible inference is an assumption that these data cannot test.
:::

::: {.callout-note collapse="true" title="Technical background: event attribution"}
Every change in a cell count must be assigned to an event. Without supplied mortality probabilities, a shrinking survivor cell is assigned to mortality and a growing survivor cell is assigned to net in-migration. This **sign attribution** leaves no residual, but it cannot identify offsetting deaths and migration. With supplied mortality probabilities, deaths are calculated first and migration becomes the signed residual of the population-balancing identity.
:::

## Prepare the outcome model and demographic inputs

All waves cover ages 18--89+. We pool ages 81 and older into a common open outcome cell. A common open cell prevents respondents from aging out of the observed range.

```{r}
gss_all <- gss_homosex[,
  .(age = pmin(age, 81), period = year, sex, y = homosex, wtssps)
]
```

We model acceptance as an additive smooth function of age and period, with a term for sex. We allow up to 15 basis functions for each smooth. This flexibility captures the sharp change in the final waves without materially changing the full-period split. The outcome model supplies predicted acceptance for each cell; it does not determine the number of people in each cell.

```{r}
model <- mgcv::gam(
  y ~ s(age, k = 15) + s(period, k = 15) + sex,
  data = gss_all,
  weights = wtssps,
  method = "REML"
)
```

The package also includes two demographic inputs:

-   `wpp_us`: U.S. population estimates by age and sex from the United Nations World Population Prospects [@unwpp2024].
-   `mortality_us`: U.S. central death rates by year, age, and sex from the Human Mortality Database [@hmd2025].

Only relative population structure matters. We therefore rescale each population wave to a tractable total. Ages 81--88 and the terminal 89+ group remain separate for demographic calculations. Their outcome predictions use the survey's 81+ cell.

```{r}
data(wpp_us)
data(mortality_us)

survey_years <- sort(unique(gss_all$period))
pop <- wpp_us[period %in% survey_years]
scale_total <- round(mean(gss_all[, .N, by = period]$N))
pop[, n := n / sum(n) * scale_total, by = period]

# decompose_aggregated() requires annual death probabilities rather than
# central death rates.
mort <- mortality_us[,
  .(period = year, age, sex, prob = 1 - exp(-death_rate))
]
```

::: {.callout-note collapse="true" title="Technical background: open age groups"}
The survey outcome model has an open 81+ cell. The population and mortality frames retain their constituent ages through a terminal 89+ group. The method uses those constituent ages for aging and mortality, clamps their outcome predictions to age 81, and reports their contributions in the 81+ outcome cell. The terminal ages in the population and mortality frames must describe the same open group.
:::

::: {.callout-note collapse="true" title="Technical background: survey weights"}
When no population frame is supplied, survey weights are normalized within each period to sum to the sample size. Rounded cell totals then preserve relative survey structure while keeping the simulation tractable. They are not population counts. A population frame is preferred when suitable estimates are available.
:::

## Build the decomposition in three stages

The stages below hold the outcome model and sex cells constant. They change only the information used to derive demographic events.

### 1. Survey counts

The simplest specification derives cell counts from the survey.

```{r}
set.seed(42)
survey_result <- decompose_aggregated(
  gss_all, model,
  cells = "sex", weight = "wtssps"
)
print(survey_result, detailed = FALSE)
```

Survey cell sizes fluctuate across waves. The function must route that fluctuation to mortality or in-migration. These detailed components must not be read as demographic estimates.

### 2. Population frame

The `population` argument lets the survey supply the outcome model while the external frame supplies the cell counts.

```{r}
set.seed(42)
population_result <- decompose_aggregated(
  gss_all, model,
  cells = "sex", weight = "wtssps",
  population = pop
)
print(population_result, detailed = FALSE)
```

The smoother population structure removes most survey cell-size noise. Net in-migration becomes small, but the mortality estimate still uses sign attribution.

### 3. Population frame and mortality probabilities

Supplied mortality probabilities determine deaths directly. Migration becomes the signed residual between estimated cohort survival and the next population count.

```{r}
set.seed(42)
mortality_result <- decompose_aggregated(
  gss_all, model,
  cells = "sex", weight = "wtssps",
  population = pop, mortality = mort
)
print(mortality_result, detailed = FALSE)
```

This specification uses the most complete demographic information available here. Its out-migration term is not a direct migration estimate. It combines net migration with inconsistency between the population estimates and mortality-based survival.

```{r}
#| code-fold: true
#| code-summary: "Show table code"
summarize_result <- function(x, specification) {
  s <- x$summary
  total <- s[.N, modeled_mean] - s[1, modeled_mean]
  intraindividual <- s[, sum(intraindividual, na.rm = TRUE)]
  mortality <- s[, sum(mortality, na.rm = TRUE)]
  coming_of_age <- s[, sum(coming_of_age, na.rm = TRUE)]
  inmigration <- s[, sum(inmigration, na.rm = TRUE)]
  outmigration <- s[, sum(outmigration, na.rm = TRUE)]

  data.frame(
    Specification = specification,
    `Total change` = total,
    `Intraindividual` = intraindividual,
    `Turnover` = mortality + coming_of_age + inmigration + outmigration,
    `Mortality` = mortality,
    `Coming-of-age` = coming_of_age,
    `In-migration` = inmigration,
    `Out-migration` = outmigration,
    check.names = FALSE
  )
}

comparison <- rbind(
  summarize_result(survey_result, "Survey counts"),
  summarize_result(population_result, "Population frame"),
  summarize_result(mortality_result, "Population frame + mortality")
)
knitr::kable(comparison, digits = 3)
```

## Substantive findings

The preferred specification attributes about 58% of the modeled increase to intraindividual change and 42% to population turnover. The broad split changes little across the three specifications. In contrast, the allocation of turnover between mortality and migration changes materially when demographic information is added.

```{r}
plot(mortality_result)
```

### Three phases of social change

The cumulative result contains three descriptive phases: two long periods with distinct patterns, followed by a short provisional reversal. The table summarizes these selected intervals.

```{r}
#| code-fold: true
#| code-summary: "Show table code"
summarize_period <- function(x, start, end) {
  s <- x$summary
  changes <- s[period > start & period <= end]

  data.frame(
    Period = paste(start, end, sep = "--"),
    `Modeled change` =
      s[period == end, modeled_mean] - s[period == start, modeled_mean],
    `Intraindividual change` =
      changes[, sum(intraindividual, na.rm = TRUE)],
    `Population turnover` = changes[, sum(
      mortality + outmigration + coming_of_age + inmigration,
      na.rm = TRUE
    )],
    check.names = FALSE
  )
}

period_comparison <- rbind(
  summarize_period(mortality_result, 1973, 1988),
  summarize_period(mortality_result, 1988, 2021),
  summarize_period(mortality_result, 2021, 2024)
)
knitr::kable(period_comparison, digits = 3)
```

From 1973 through 1988, modeled acceptance declined slightly. Intraindividual change contributed −0.070, while population turnover contributed +0.054. Turnover was already moving the population toward greater acceptance, but change within the surviving population moved in the opposite direction. The aggregate trend therefore conceals two opposing social processes.

This early pattern matches the historical sequence described by Ekstam [-@ekstam2021]: advances associated with gay-rights mobilization met organized conservative countermobilization in the 1970s, followed by a more hostile climate during the HIV/AIDS crisis of the 1980s. Research documents both the role of the religious right in shaping countermobilization [@fetner2008] and changing attitudes during the epidemic [@ruelcampbell2006]. These accounts support the periodization, but they do not establish the cause of the negative component.

From 1988 through 2021, modeled acceptance rose by 0.489. Intraindividual change contributed +0.378, compared with +0.111 from population turnover. Turnover remained positive, but change within the existing population became the main source of rising acceptance. This pattern is consistent with a shift from replacement offsetting resistance to broader cultural diffusion across the population.

Proposed mechanisms for this diffusion include gay-rights movement campaigns [@epstein1999], changing elite messages [@brewer2003], greater media visibility and parasocial contact [@schiappa2006], and declining religious opposition [@sherkat2011]. The decomposition does not distinguish among these mechanisms.

#### Post-2021 reversal?

From 2021 through 2024, modeled acceptance fell by 0.056. Intraindividual change contributed −0.065, while population turnover contributed +0.009. This resembles the early period: positive turnover partly offsets negative intraindividual change.

The decomposition adds a demographic qualification to accounts of recent anti-LGBTQ+ backlash. Population turnover continued to move acceptance upward after 2021, but modeled change within the surviving population moved it downward by a larger amount. Once coming-of-age, mortality, and net migration are accounted for, the most plausible substantive interpretation is that people already in the adult population became less accepting. A panel would observe such changes directly; this decomposition supports the inference indirectly through aggregate change and demographic accounting.

The timing coincides with organized countermobilization. Jones [-@jones2024] identifies 543 state anti-LGBTIQ+ bills introduced from 2018 through 2022, commonly framed through parental, religious, and women's rights. McDowell and Ward [-@mcdowellward2023] show how evangelical churchgoers framed conservative Christians as marginalized to justify religious exemptions that permit denial of services to LGBTQ+ people. These studies establish a relevant political context, but they do not show that countermobilization caused the change estimated here. Much of the recent campaign targeted transgender people and youth, so its connection to attitudes toward homosexuality remains indirect.

This short reversal does not establish a new sociological period. It contains only two transitions, and its boundary follows the observed 2021 peak. Survey modes also changed after 2021. A sensitivity check using the comparable 2022 and 2024 waves finds declines within each mode and after standardizing the mode distribution. Changing mode composition therefore does not explain the 2022--2024 decline, although other measurement effects remain possible.

::: {.callout-note collapse="true" title="Mode-sensitivity check"}
The 2021 round had no in-person interviews. The 2022 and 2024 rounds included the same four mode categories, which permits a more direct comparison.

```{r}
mode_comparison <- gss_homosex[
  year %in% c(2022, 2024) & !is.na(mode),
  .(
    acceptance = weighted.mean(homosex, wtssps),
    mode_weight = sum(wtssps)
  ),
  by = .(year, mode)
]

mode_mix_2022 <- mode_comparison[year == 2022, .(
  mode,
  share = mode_weight / sum(mode_weight)
)]
mode_standardized <- mode_comparison[mode_mix_2022, on = "mode"][, .(
  acceptance = sum(acceptance * share)
), by = year]

mode_table <- data.table::dcast(
  mode_comparison, mode ~ year, value.var = "acceptance"
)
mode_table <- rbind(
  mode_table[, .(Mode = mode, `2022` = `2022`, `2024` = `2024`)],
  data.table::data.table(
    Mode = "Standardized to 2022 mode mix",
    `2022` = mode_standardized[year == 2022, acceptance],
    `2024` = mode_standardized[year == 2024, acceptance]
  )
)
knitr::kable(mode_table, digits = 3)
```

Acceptance declined within every interview mode. Holding the mode distribution at its 2022 composition increases the estimated decline from 0.062 to 0.082. Mode is a measurement condition rather than a demographic event, so it is not passed to the decomposition as a population cell.
:::

### Contributions by sex

The decomposition is additive across the supplied cells. Passing `covariate = "sex"` attributes each component to the sex cells that produced it. These are contributions to aggregate change, not sex-specific rates of change.

```{r}
print(mortality_result, detailed = FALSE, covariate = "sex")
plot(mortality_result, covariate = "sex")
```

Women and men contribute similar amounts to the total increase. Mortality and residual out-migration differ in their point estimates, but these values do not identify sex differences in emigration. The migration component remains a residual difference between two demographic sources.

## Uncertainty

Passing `R > 0` adds bootstrap draws and empirical 95% intervals. Each replicate uses a new random event ordering and a reweighted refit of the outcome model. Here, `R = 100` keeps runtime manageable; its tail estimates are illustrative.

```{r}
set.seed(42)
bootstrap_result <- decompose_aggregated(
  gss_all, model,
  cells = "sex", weight = "wtssps",
  population = pop, mortality = mort,
  R = 100, seed = 42
)
print(bootstrap_result, detailed = FALSE)
plot(bootstrap_result)
```

The intervals quantify combined outcome-model and event-ordering uncertainty. They do not include uncertainty in the population frame or mortality probabilities. The reweighted bootstrap also does not reproduce the GSS sample design or resample primary sampling units. It must not be interpreted as a full design-based uncertainty estimate.

Sex-specific contributions and intervals are:

```{r}
print(bootstrap_result, detailed = FALSE, covariate = "sex")
plot(bootstrap_result, covariate = "sex")
```

Marginal intervals for women and men do not test their difference. A sex contrast must be calculated within each bootstrap replicate.

::: {.callout-note collapse="true" title="Technical background: event ordering"}
The timing and order of demographic events between survey waves are not observed. The method assigns events to evenly spaced positions in random order. With `R = 0`, the point estimate uses one ordering. With `R > 0`, it averages over `R` orderings and pairs each ordering with one model refit.
:::

## Comparison with CR-IC decomposition

The classical cohort-replacement and intracohort-change framework provides a useful benchmark. `cr_ic()` implements algebraic decomposition (AD), linear decomposition (LD), and two model-based improvements (AD+ and Model). For a detailed explanation and a replication of Firebaugh [-@firebaugh1989], see the [Replications: Firebaugh](replicating_firebaugh.html) vignette.

Using the same GSS item through 2016, Ekstam [-@ekstam2021] shows that estimated linear age, period, and cohort trends change substantially with the identifying constraints imposed by an APC model. `decompose_aggregated()` does not fit a conventional APC model. It defines components through modeled outcome trajectories and demographic events, with the assumptions described above.

Using all waves, AD+ attributes about 60% of the increase to intracohort change and 40% to cohort replacement. The Model method gives 46% and 54%. These estimates bracket the broad split from `decompose_aggregated()`.

```{r}
form <- homosex ~ as.factor(year) + as.factor(cohort)
cric_all <- cr_ic(
  gss_homosex,
  homosex ~ year + cohort,
  weight = "wtssps", model = form
)
cric_all
plot(cric_all)
```

Using only the first and last survey years changes the decomposition. AD+ attributes about 59% to cohort replacement and 41% to intracohort change; the Model method gives 55% and 45%. This sensitivity shows why the selected waves and functional form must be reported.

```{r}
form <- homosex ~ as.factor(year) + splines::bs(cohort, 10)
cric_endpoints <- cr_ic(
  gss_homosex[year %in% c(min(year), max(year))],
  homosex ~ year + cohort,
  weight = "wtssps", model = form
)
cric_endpoints
```

CR-IC summarizes change as intracohort change and cohort replacement. `decompose_aggregated()` instead tracks coming-of-age, mortality, and residual migration and shows when each component accumulated.

## References

::: {#refs}
:::
